---
title: "HW2 Optimization"
author: "Jing Li"
date: "October 21, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 2.1 Matrix Representation

Using R, we can generate a random n×n adjacency matrix as follows. In the example below, each edge has a 20% chance of occuring. 

```{r}
n <- 50
d <- runif(n*n)
d[d < 0.8] <- NA
d <- matrix(d, nrow = n, ncol = n)
diag(d) <- NA
d[upper.tri(d)] = t(d)[upper.tri(d)]
```


## 2.2 Adjacency Matrix to Adjacency List

Write a function to convert adjacency matrix to an adjcency list called ds

```{r}
AdjMatrix2List <- function(d){
  ds <- matrix(ncol = 3,dimnames = list(NULL,c("head","tail","weight")))
  for (i in 1:nrow(d)){
    for (j in 1:nrow(d)){
      if(!is.na(d[i,j])){
        ds <- rbind(ds,c(i,j,d[i,j]))
      }
    }
  }
  ds <- ds[-1,]
  return(ds)
}
head(AdjMatrix2List(d),50)
```


## 3 Euclidean Minimum Spanning Tree

Create a euclidean minimum tree on a set of random (x,y) coordinates. We want to
build and visualize a network that connects all of the coordinates at minimum 
cost we can generate a random set of 50 coordinates 

```{r}
n <- 50
x <- round(runif(n)*1000)
y <- round(runif(n)*1000)
plot(x,y,pch=16)
```


## 3.1 Adjacency Matrix 

Create an adjacency matrix d by calculating the euclidean distance between every 
pair of points

```{r}
eudist <- function(x,y){
  distance <- c()
  for (i in 1:n){
    for(j in 1:n){
      distance <- c(distance, sqrt((x[j]-x[i])^2+(y[j]-y[i])^2))
    }
  }
  distance <- matrix(distance, nrow = n, ncol = n)
  return(distance)
}

d <- eudist(x,y)
head(d,10)
```


## 3.2 Adjacency List

Create an adjacency list from the distance adjacency matrix

```{r}
ds <- AdjMatrix2List(d)
head(ds,50)
```


## 3.3 Calculate the Minimum Spanning Tree using Kruskal or Prim

```{r}
require(optrees)
ds.mst <- msTreePrim(1:n, ds)
str(ds.mst)
```


## 3.4 Produce a Plot Visulizing the Minimum Spanning Tree

Using the segments function, draw straight line segments on the plot (x,y)
corresponding to the edge in the minimum spanning tree. ept1 is the source code
and ept2 is the destination node. ignore the weights

```{r}
plot(x,y)
xstart <- c()
ystart <- c()
xend <- c()
yend <- c()

plot.mst <- function(ds.mst){
  for(i in 1:nrow(ds.mst$tree.arcs)){
    xstart[i] <- x[ds.mst$tree.arcs[i,][1]]
    ystart[i] <- y[ds.mst$tree.arcs[i,][1]]
    xend[i] <- x[ds.mst$tree.arcs[i,][2]]
    yend[i] <- y[ds.mst$tree.arcs[i,][2]]
  }
  return(segments(xstart,ystart,xend,yend))
}
plot.mst(ds.mst)
```


## 4 Hostile Agents

An intelligence service has n agents in a non-friendly country. Each agent knows 
some of the other agents and has in place procedures for arranging a rendezvous 
with anyone he or she knows. For each such possible rendezvous, say between 
agents i and j, any message passed between these agents will fall into hostile 
hands with a certain probability pij. The government wants to transmit a 
confidential message among all the agents while maximizing the total probability 
that no message is intercepted


## 4.1 Specify This Problem As a Graph Optimization Problem

There are n nodes in a graph. Each nodes has some edges connecting to other 
nodes. The weight for each edges is a random number between 0 and 1. The task is 
to specify a starting node then construct a minimum spanning tree that connect 
all nodes at a minimum cost (overall probability of the intelligence being 
captured by enemy is minimized) 


## 4.2 Structure Inputs to This Problem

We need to initialize a partition structure which has two sets, "key" and "pred"
For each vertex v from V, set key[v] = infinity and set pred[v] = -1. We also
want to specify key[0] = 0 which indicates our starting node. We then write a 
priority queue PQ. 

The essential "inputs" in this algorithm are E which represents vectors 
containing all edges and V which represents vectors containing all vertex.


## 4.3 Which Algorithm to Use

The Kruskal algorithm is preferred. Although, both the Prim algorithm and the 
Kruskal algorithm will construct a minimum spanning tree; and, both algorithms 
will capture the same of amount of information about the edges (in our case are 
the individual probabilities that enemy will capture the intelligence) 
However since we are dealing with probabilities, we would want to multiply 
probabilities along each edge in order to deliver the intelligence from a 
starting node to a ending node. 

Think this as partitioning sets. There are more intermediate parent nodes 
between a root node and a individual node in a minimum spanning tree that is 
produced by the Kruskal algorithm than one produced by the Prim algorithm. 


## 4.4 What is Computational Efficiency 

The running time for Dijkstra's algorithm is O(E lg E)


## 5 Project Scheduling

Consider a task A which must be completed before some task B. We can represent 
all such tasks in a directed acyclic graph (DAG), such that if A happens before 
B, we include a directed edge from A to B. 


## 5.1 Matrix Representation

Represent this scheduling network in R. Associate a completion time with each 
node.

```{r}
adj_matrix1 <- matrix(NA,nrow = 10,ncol = 10,dimnames = list(c('A','B','C','D','E','F','G','H','I','J'),c('A','B','C','D','E','F','G',
                                                                                                          'H','I','J')))
adj_matrix1[1,2] <- 90
adj_matrix1[1,6] <- 90
adj_matrix1[1,9] <- 90
adj_matrix1[2,3] <- 15
adj_matrix1[3,7] <- 5
adj_matrix1[4,5] <- 20
adj_matrix1[4,8] <- 20
adj_matrix1[4,10] <- 20
adj_matrix1[6,7] <- 25
adj_matrix1[7,4] <- 14
adj_matrix1[9,10] <- 30
adj_matrix1
```


## 5.2 Enter, the Management

1. The earliest completion date for the project if it commences on Nov 1, 2017. 
2. The earliest and latest start times for each task which will not alter the 
   completion date. 
3. The earliest and latest ﬁnish times for each activity that will not alter the            completion date. 
4. The list of tasks which must adhere to a rigid schedule so as not to delay 
   the overall project, as well as a list of those tasks which have more 
   ﬂexibility. 


## 5.4 Earliest Start Times (ES)

To calculate the earliest start times for each node, you will need to solve the 
longest path problem, starting at source node A, with negative edge weights. The distances output from the algorithm are your earliest start times. 

```{r}
require(optrees)
adj_matrix1 <- adj_matrix1 * -1
adj_list1 <- AdjMatrix2List(adj_matrix1)
ES <- spTreeBellmanFord(1:10, adj_list1,source.node = 1, directed = T)$distances*-1
ES
```


## 5.6 Earliest Finish Times

EF = ES + node duration

```{r}
s.nodes <- c(90,15,5,20,21,25,14,28,30,45)
EF <- ES + s.nodes
EF
```


## 5.7 Latest Finish Time

The latest ﬁnish times are computed on the transpose of the graph. That is, for 
each edge from A to B, we reverse its direction and instead have an edge from B 
to A. we will have to rebuild your edge weights. We are still looking for ﬁnish 
times along the (longest) critical path, so again, we ﬂip the edge weights to 
negative, and solve the shortest path problem.

We can add an 11th node — a dummy node with duration 0 which has outgoing edges 
to E, H, and J.

```{r}
require(optrees)
adj_matrix2 <- matrix(NA,nrow = 11,ncol = 11,dimnames = list(c('A','B','C','D','E','F','G','H','I','J',"K"),c('A','B','C','D','E','F',
                                                                                                              'G','H','I','J',"K")))

adj_matrix2[2,1] <- 15
adj_matrix2[6,1] <- 25
adj_matrix2[9,1] <- 30
adj_matrix2[3,2] <- 5
adj_matrix2[7,3] <- 14
adj_matrix2[5,4] <- 21
adj_matrix2[8,4] <- 28
adj_matrix2[10,4] <- 45
adj_matrix2[7,6] <- 14
adj_matrix2[4,7] <- 20
adj_matrix2[10,9] <- 45
adj_matrix2[11,8] <- 0
adj_matrix2[11,5] <- 0
adj_matrix2[11,10] <- 0
adj_matrix2

adj_matrix2 <- adj_matrix2 * -1
adj_list2 <- AdjMatrix2List(adj_matrix2)
LF <- spTreeBellmanFord(1:11, adj_list2,source.node = 11, directed = T)$
  distances * -1
LF <- EF[length(EF)]-LF[-length(LF)]
LF
```


## 5.8 Latest Start Times

LS = LF - node duration

```{r}
LS <- LF - s.nodes
LS
```


## 5.9 Slack

Once we have computed the above quantities, the slack along the critical path is
either LF − EF or LS − ES. 

1. Which tasks have scheduling ﬂexibility? (positive slack) 
2. Which tasks are on the critical path? (slack=0)

```{r}
s.labels <- c('A','B','C','D','E','F','G','H','I','J')
Critical_Path <- c()
Flexibility <- c()
for(i in 1:10){
  if((LF[i] - EF[i]) == 0)
    Critical_Path[i] <- s.labels[i]
  else
    Flexibility[i] <- s.labels[i]
}
Flexibility[10] <- NA
final_table <- data.frame(Critical_Path,Flexibility,ES,EF,LS,LF,row.names = 
                            s.labels)
final_table 
```
